Supplementary Information from:\\ \textit{Reconstructing the last known movements of one of Nature's giants}

### 2018-01-31 manuscript/revision/blue-whale-supplementary-v2.tex
change Fig to Figure

\subsection*{CHARACTERISATION OF THE MEASURED BALEEN}
The baleen plate yielded 97 discrete samples of baleen. 
Across the baleen sample, $\delta^{15}$N values show clear cycles with periods of relatively high $\delta^{15}$N values interspersed by relatively short periods where $\delta^{15}$N values are c.1\text{\textperthousand} lower. 
Fourier analysis (periodogram function in the R package TSA\cite{Chan:2012aa}) revealed a strong periodic repetition with a 13.3cm periodicity. 
The oldest (most distal) 35cm of the baleen plate  record relatively positive and constant $\delta^{13}$C values, associated with a weakening of the periodic fluctuation seen in $\delta^{15}$N values, and an overall reduction in $\delta^{15}$N values. 
$\delta^{13}$C values also show repeated fluctuations in the middle 50cm of the baleen plate (behavioural phase two), with a similar periodicity of 13.5cm (Fig. \ref{figs1}). 
A distinct change in the isotopic pattern along the baleen plate occurs at around 84cm from the distal end in behavioural phase two, with a transition to uniform, relatively low $\delta^{13}$C values followed by an abrupt transition to positive $\delta^{13}$C values and a decline in $\delta^{13}$C values in the most recent (most gingival) 3cm of the record.

@@

Cross-correlation analysis demonstrates a strong negative covariance between $\delta^{13}$C and $\delta^{15}$N  values within behavioural phase two (Fig. \ref{figs2}), but no relationship between $\delta^{13}$C and $\delta^{15}$N values exists during behavioural phase one.

@@

Simulated $\delta^{15}$N POM values are relatively positive in the northeast Atlantic north of c.60$^{\circ}$N, and relatively negative in the central and southern North Atlantic, associated with high levels of nitrogen fixation (Fig. \ref{figs3}). 
Annual average $\delta^{13}$C POM values largely vary with latitude, with more negative values in more northerly regions. 
In the central North Atlantic, $\delta^{13}$C POM values are relatively positive in the west, reflecting warm gulf stream waters (Fig. \ref{figs3}). 
The temporally coincident periodic, inverse fluctuations in $\delta^{13}$C and $\delta^{15}$N values in the second phase of the baleen isotopic record are therefore consistent with migration between waters north of 60$^{\circ}$N and more southerly southern waters in the North Atlantic, with northerly occupancy in summer months and southerly occupancy in winter months. 
Incorporation of carbon and nitrogen with isotopic ratios consistent with southerly latitudes during winter months indicates continued feeding during migration and winter residency.

@@

\subsection*{ISOTOPE MODEL SIMULATIONS}
We initially simulated temporal variations in baseline (phytoplankton) isotopic compositions that would be encountered by whales foraging within broad geographic regions (west UK/Irish shelf, Norwegian/Barents Sea, Canaries/west Azores, Mid Atlantic ridge, Cape Verde/Mauritanean upwelling area; Fig. \ref{figs4}). 
Strong seasonal dynamics in $\delta^{13}$C values are evident in all northerly regions, characterized by a rapid increase to annual maximum $\delta^{13}$C values associated with the onset of the spring phytoplankton bloom, followed by more gradual decline in $\delta^{13}$C values towards minima in winter conditions (Fig. \ref{figs4}). 
In temperate latitudes around the British Isles, seasonal dynamic cycles are present, but dampened (Fig. \ref{figs4}), whereas in subtropical regions exemplified by the Canaries, Azores and particularly Mauritanean upwelling areas, $\delta^{13}$C values are relatively positive and constant throughout the year (Fig. \ref{figs4}).

@@

\subsubsection*{Matching isotope model simulations with whale isotopes in behavioural phase one}
Behavioural phase one is characterised by approximately 24 months of relatively positive and constant $\delta^{13}$C values. 
$\delta^{15}$N values in this phase are relatively low and the isotopic cyclicity is absent (Fig. \ref{figs1}). 
Positive and temporally constant $\delta^{13}$C values are characteristic of subtropical regions particularly upwelling areas around Mauritania, Cape Verde and the Azores region.

\subsubsection*{Matching isotope model simulations with whale isotopes in behavioural phase two}
Temporal dynamics in $\delta^{13}$C values simulated in northerly latitudes cannot be reproduced from simulated temporal trends in $\delta^{13}$C values in krill within static areas (Fig. \ref{figs4}, Fig. 3). 
Adding seasonal north-south migrations within mid-high latitude regions to foraging models, closely matches the measured profile, including $\delta^{13}$C minima preceding the June bloom (Fig. \ref{figs4}, Fig. 1).

Accordingly we modeled whale movements allowing three years of residency in warm waters of the Azores or Cape Verde/Mauritania region followed by a further 3.5 years of seasonal north-south migration in the northeast northern Atlantic, followed by residency in subtropical waters in 1890 and rapid migration returning to more northerly waters in the spring of 1891. 
The resulting simulated profiles closely match the measured baleen plate, capturing all major temporal fluctuations (Fig. 3). 
Blue whales have a gestation period of 10-12 months, and calves suckle for approximately 6-7 months \cite{handbook}. 
The proposed timescale of pregnancy around early spring 1889, birth in spring 1890, lactation through summer and autumn 1890 and return to northern feeding grounds in early 1891 is therefore fully consistent with the assumed reproductive ecology of blue whales.

\subsection*{AGENT-BASED WHALE MOVEMENT MODEL}
We simulated the likely isotopic expression associated with feeding and movement behaviours by building an agent-based model sampling $\delta^{13}$C POM values predicted by the isotopic extension to the NEMO-MEDUSA model\cite{magozzi2017using,yool2013medusa}, where movement behavior is influenced by sea surface temperature and phytoplankton biomass estimates provided by NEMO-MEDUSA\cite{yool2013medusa}, and bathymetry (Fig. \ref{figs5}).

@@

\centering
\begin{table}
  \begin{tabular}{|p{8cm}|c|c|c|} 
    \hline
    & \multicolumn{3}{|c|}{\textbf{Behavioural state}} \\
    \hline
    \textbf{Parameters} & \textbf{Foraging} & \textbf{Migrating north} & \textbf{Migrating south}\\
    \hline
    Behavioural phase one (1095 days) & Jan - Dec & NA & NA\\
    \hline
    Behavioural phase two (1278 days) & June - Oct & April - May & Nov - March\\
    \hline
    Pregnancy, nursing, return (447 days) & Jan - Dec & NA & NA\\
    \hline
    Post nursing migration (100 days) & NA & Jan - June & NA\\
    \hline
    Minimum sea surface temperature$^*$ ($^{\circ}$C) & 3 & 3 & 3\\
    \hline
    Maximum sea surface temperature ($^{\circ}$C) & 25 & 25 & 25\\
    \hline
    Maximum daily distance traveled (km)$^{**}$ & $N(\mu=50, \sigma=25)$ & $N(\mu=150, \sigma=75)$ & $N(\mu=150, \sigma=75)$\\
    \hline
    Plankton biomass threshold for remaining to forage (units) & 50 & NA & NA\\
    \hline
  \end{tabular}
  \caption{Model parameters for agent-based model of whale movement. 
  See Fig. \ref{figs5} and text for more details. 
  $^*$During nursing minimum sea surface temperature is set to 18$^{\circ}$C. 
  $^{**}$During behavioural phase one, mean maximum daily distance is set to 20km.}
  \label{tables1}
\end{table}

@@

  \begin{figure}[!htbp]
    \centering
      \includegraphics[width=\linewidth]{figures/Figure-S5-r2.png}
      \caption{Histogram of $r^2$ values from linear regressions comparing model simulation $\delta^{13}$C values to those of the real data (Fig. 1).} % check legend
      \label{figs6}
  \end{figure}

>>>>

\subsection*{CHARACTERISATION OF THE MEASURED BALEEN}
The baleen plate yielded 97 discrete samples of baleen. 
Across the baleen sample, $\delta^{15}$N values show clear cycles with periods of relatively high $\delta^{15}$N values interspersed by relatively short periods where $\delta^{15}$N values are c.1\text{\textperthousand} lower. 
Fourier analysis (periodogram function in the R package TSA\cite{Chan:2012aa}) revealed a strong periodic repetition with a 13.3cm periodicity. 
The oldest (most distal) 35cm of the baleen plate  record relatively positive and constant $\delta^{13}$C values, associated with a weakening of the periodic fluctuation seen in $\delta^{15}$N values, and an overall reduction in $\delta^{15}$N values. 
$\delta^{13}$C values also show repeated fluctuations in the middle 50cm of the baleen plate (behavioural phase two), with a similar periodicity of 13.5cm (Figure \ref{figs1}). 
A distinct change in the isotopic pattern along the baleen plate occurs at around 84cm from the distal end in behavioural phase two, with a transition to uniform, relatively low $\delta^{13}$C values followed by an abrupt transition to positive $\delta^{13}$C values and a decline in $\delta^{13}$C values in the most recent (most gingival) 3cm of the record.

@@

Cross-correlation analysis demonstrates a strong negative covariance between $\delta^{13}$C and $\delta^{15}$N  values within behavioural phase two (Figure \ref{figs2}), but no relationship between $\delta^{13}$C and $\delta^{15}$N values exists during behavioural phase one.

@@

Simulated $\delta^{15}$N POM values are relatively positive in the northeast Atlantic north of c.60$^{\circ}$N, and relatively negative in the central and southern North Atlantic, associated with high levels of nitrogen fixation (Figure \ref{figs3}). 
Annual average $\delta^{13}$C POM values largely vary with latitude, with more negative values in more northerly regions. 
In the central North Atlantic, $\delta^{13}$C POM values are relatively positive in the west, reflecting warm gulf stream waters (Figure \ref{figs3}). 
The temporally coincident periodic, inverse fluctuations in $\delta^{13}$C and $\delta^{15}$N values in the second phase of the baleen isotopic record are therefore consistent with migration between waters north of 60$^{\circ}$N and more southerly southern waters in the North Atlantic, with northerly occupancy in summer months and southerly occupancy in winter months. 
Incorporation of carbon and nitrogen with isotopic ratios consistent with southerly latitudes during winter months indicates continued feeding during migration and winter residency.

@@

\subsection*{ISOTOPE MODEL SIMULATIONS}
We initially simulated temporal variations in baseline (phytoplankton) isotopic compositions that would be encountered by whales foraging within broad geographic regions (west UK/Irish shelf, Norwegian/Barents Sea, Canaries/west Azores, Mid Atlantic ridge, Cape Verde/Mauritanean upwelling area; Figure \ref{figs4}). 
Strong seasonal dynamics in $\delta^{13}$C values are evident in all northerly regions, characterized by a rapid increase to annual maximum $\delta^{13}$C values associated with the onset of the spring phytoplankton bloom, followed by more gradual decline in $\delta^{13}$C values towards minima in winter conditions (Figure \ref{figs4}). 
In temperate latitudes around the British Isles, seasonal dynamic cycles are present, but dampened (Figure \ref{figs4}), whereas in subtropical regions exemplified by the Canaries, Azores and particularly Mauritanean upwelling areas, $\delta^{13}$C values are relatively positive and constant throughout the year (Figure \ref{figs4}).

@@

\subsubsection*{Matching isotope model simulations with whale isotopes in behavioural phase one}
Behavioural phase one is characterised by approximately 24 months of relatively positive and constant $\delta^{13}$C values. 
$\delta^{15}$N values in this phase are relatively low and the isotopic cyclicity is absent (Figure \ref{figs1}). 
Positive and temporally constant $\delta^{13}$C values are characteristic of subtropical regions particularly upwelling areas around Mauritania, Cape Verde and the Azores region.

\subsubsection*{Matching isotope model simulations with whale isotopes in behavioural phase two}
Temporal dynamics in $\delta^{13}$C values simulated in northerly latitudes cannot be reproduced from simulated temporal trends in $\delta^{13}$C values in krill within static areas (Figure \ref{figs4}, Figure 3). 
Adding seasonal north-south migrations within mid-high latitude regions to foraging models, closely matches the measured profile, including $\delta^{13}$C minima preceding the June bloom (Figure \ref{figs4}, Figure 1).

Accordingly we modeled whale movements allowing three years of residency in warm waters of the Azores or Cape Verde/Mauritania region followed by a further 3.5 years of seasonal north-south migration in the northeast northern Atlantic, followed by residency in subtropical waters in 1890 and rapid migration returning to more northerly waters in the spring of 1891. 
The resulting simulated profiles closely match the measured baleen plate, capturing all major temporal fluctuations (Figure 3). 
Blue whales have a gestation period of 10-12 months, and calves suckle for approximately 6-7 months \cite{handbook}. 
The proposed timescale of pregnancy around early spring 1889, birth in spring 1890, lactation through summer and autumn 1890 and return to northern feeding grounds in early 1891 is therefore fully consistent with the assumed reproductive ecology of blue whales.

\subsection*{AGENT-BASED WHALE MOVEMENT MODEL}
We simulated the likely isotopic expression associated with feeding and movement behaviours by building an agent-based model sampling $\delta^{13}$C POM values predicted by the isotopic extension to the NEMO-MEDUSA model\cite{magozzi2017using,yool2013medusa}, where movement behavior is influenced by sea surface temperature and phytoplankton biomass estimates provided by NEMO-MEDUSA\cite{yool2013medusa}, and bathymetry (Figure \ref{figs5}).

@@

\centering
\begin{table}
  \begin{tabular}{|p{8cm}|c|c|c|} 
    \hline
    & \multicolumn{3}{|c|}{\textbf{Behavioural state}} \\
    \hline
    \textbf{Parameters} & \textbf{Foraging} & \textbf{Migrating north} & \textbf{Migrating south}\\
    \hline
    Behavioural phase one (1095 days) & Jan - Dec & NA & NA\\
    \hline
    Behavioural phase two (1278 days) & June - Oct & April - May & Nov - March\\
    \hline
    Pregnancy, nursing, return (447 days) & Jan - Dec & NA & NA\\
    \hline
    Post nursing migration (100 days) & NA & Jan - June & NA\\
    \hline
    Minimum sea surface temperature$^*$ ($^{\circ}$C) & 3 & 3 & 3\\
    \hline
    Maximum sea surface temperature ($^{\circ}$C) & 25 & 25 & 25\\
    \hline
    Maximum daily distance traveled (km)$^{**}$ & $N(\mu=50, \sigma=25)$ & $N(\mu=150, \sigma=75)$ & $N(\mu=150, \sigma=75)$\\
    \hline
    Plankton biomass threshold for remaining to forage (units) & 50 & NA & NA\\
    \hline
  \end{tabular}
  \caption{Model parameters for agent-based model of whale movement. 
  See Figure \ref{figs5} and text for more details. 
  $^*$During nursing minimum sea surface temperature is set to 18$^{\circ}$C. 
  $^{**}$During behavioural phase one, mean maximum daily distance is set to 20km.}
  \label{tables1}
\end{table}

@@

  \begin{figure}[!htbp]
    \centering
      \includegraphics[width=\linewidth]{figures/Figure-S5-r2.png}
      \caption{Histogram of $r^2$ values from linear regressions comparing model simulation $\delta^{13}$C values to those of the real data (Figure 1).} % check legend
      \label{figs6}
  \end{figure}
